Learning how to act: making good decisions with machine learning

### 2017-06-25 thesis/thesis_map.tex
working on chapter 4

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    punkt/.style={
           circle,
           rounded corners,
           draw=black, thick,
           text width=1em,
           minimum height=1em,
           text centered},
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered
           },
    target/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
    observedrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    latentrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered
           },
     targetrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }   
}

\newcommand{\KL}{\operatorname{KL}}
\newcommand{\simpleregret}{R_T}
\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
\newcommand{\Q}[1]{\operatorname{Q}\left\{#1\right\}}
\newcommand{\EE}{\mathbb E}
\newcommand{\EEa}{\EE_a}
\newcommand{\Pns}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
\newcommand{\Pn}[2]{\operatorname{P}\left\{#2|#1\right\}}
\newcommand{\actions}{\mathcal{A}}
\newcommand{\calA}{\mathcal A}
\newcommand{\etc}{\textit{etc}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\calP}{\mathcal P}
\newcommand{\x}{\boldsymbol{x}}
\newcommand{\Ps}{\operatorname{P}}
\newcommand{\R}{\mathbb R}
\newcommand{\Pri}[1]{\operatorname{P}_i\left\{#1\right\}}
\newcommand{\Prz}[1]{\operatorname{P}_0\left\{#1\right\}}

\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}[1]{\mathbb E\left[{#1}\right]}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}

\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
\newcommand{\cf}[2]{{#1}^{#2}}
\newcommand{\bernoulli}{\operatorname{Bernoulli}}
\newcommand{\dirac}{\operatorname{Dirac}}
\newcommand{\parents}[1]{\operatorname{\mathcal{P}a}_{#1}}
\renewcommand{\vec}[1]{\boldsymbol{#1}}

@@

@@

The rise of big data sets and powerful computers has seen an explosion in the application of machine learning. From healthcare, to entertainment and self driving cars; machine learning algorithms will transform many industries. It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge data sets heralds the end of theory (Reference) and that we may be only a short step from the Singularity, where artificial intelligence exceeds our own and then grows exponentially.

@@

The notion of causality has been widely debated in science and philosophy \cite{} but is still widely viewed as poorly defined \cite{}. This has led to a reluctance among applied researchers in many fields to refer to cauality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}. However, the magnitude, direction and even existance of an association depends on what other variables we adjust for (or include in a regression). Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via common sense if the association reported is the \emph{right one}.

@@

I do not find this distinction useful. 
We can only change the future - history is useful only as far as it tells us something about the future.
Reverse causal questions can be reposed as forward ones, when making the translation a reverse causal query will be effectively asking about many possible interventions (rather than just one). Problems highlighted as intractible in the reverse causal sense are also intractible in the forward inference form, typically because concern situations for which we do not have a sufficient number of similar instances to allow statistical reasoning. For example, the war question posed in Gelman.

A distinction between forward causal iference, what happens if we do X and reverse casual inference

@@

\subsection{Identifying when we have a causal problem}

@@

What about predicting the risk of death in patients with pneumonia? Suppose we wish to use the model to decide who should be treated in hospital and who can be sent home with antibiotics. If we assume that in hospital treatment is more effective this seems like a straightforward prediction problem. It is not. Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if we start using the model to decide whom to admit. (xxx et al) found exactly this effect. Their model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pneumonia. They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk. There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original addmition and treatment protocols. However, using it to decide on what basis to admit people could kill. The key is understanding exactly what question you are asking. In this case we are care about what happens to patients with characteristics X if we treat them according to decision rule Z.

@@

\item To decide between actions we only need to rank them (not estimate their actual effect). 
\item The predicted outcome in the absence of an intervention provides a single point. We can use this to find which problems are most serious if left alone - and prioritise those for modelling changes. 
\item Any decision we take does not significantly impact the system from which the data was drawn to make it (for repeat decision making)
\item Does acting on the result of the prediction change the predictive distribution p(y|x)? Ie change people's behaviour.
\item Ethics - ... People's viewpoint on if its OK...
\end{itemize}

@@

Causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it. To do this we need terminology to describe actions and how we anticipate the system should respond to them. Three key approaches have emerged; counterfactuals, structural equation models and causal bayesian networks. In this chapter we describe these approaches. We will examine what problems they allow us to solve, what assumptions they rely on and how they differ. We will also use them to describe the following simplified examples. The aim is to demonstrate the notations and formalisms we will need to tackle more interesting problems later on.

\vspace*{.3cm}
\begin{example}
\label{exm:ranomized_experiment}
Suppose we have developed a new drug for some illness and wish to determine how effective it is. We take a large group of patients and randomly assign half of them to a treatment group and the other half to a control group. The people in the treatment group get the drug, everyone else gets a placebo pill. The question we want to answer is does giving people the active drug improve their changes of recovery relative to giving them the placebo. We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome. 
\end{example}

\vspace*{.3cm}
\begin{example}
\label{exm:adjusting}
Suppose we wish to estimate what would happen to high school graduation rates if we made pre-school available and compulsory for all 4 year olds. We have a large cross-secional dataset on a group of 20 year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES). We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectivness of early childhood education including the landmark randomised trial, the Perry Preschool project \citep{weikart1970longitudinal}}
\end{example}

\section{Causal bayesian networks}

@@

\begin{figure}[h]
\centering
\caption{A general Bayesian network for the joint distribution over three variables. This network does not encode any conditional independencies between its variables and can thus represent any distribution over three variables.}
\label{fig:bayesnet}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]

@@

\begin{figure}
\centering
\caption{Some valid Bayesian networks for a distribution that an be factorised as $\P{X_1,X_2,X_3} = \P{X_1}\P{X_2}\P{X_3|X_2}$ (which implies $X_3 \ci X_1 | X_2$)}
\label{fig:bayesnet2}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={punkt}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={punkt}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (3) edge (2)
    (2) edge (1);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={punkt}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, below left=of 1](2){$X_{1}$};
\node[main node, below right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={punkt}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, above left=of 1](2){$X_{1}$};
\node[main node, above right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (2) edge (1)
    (3) edge (1)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\end{figure}

The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorised over the nodes and edges in the graph. There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$ (the converse is not true $G$ can have extra edges). If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;

\begin{equation}
P(X_{1}...X_{N}) = \prod_{i = 1...N}P(X_{i}|parents_{X_{i}})
\end{equation}

A causal Bayesian network is a Bayesian network in which it a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$. This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change. We need some notation to describe interventions and represent distributions over variables in the network after an intervention. In this thesis I use the do operator introduced by Pearl \citep{Pearl2000}.

\begin{definition}{The do-notation}
\begin{itemize}
\item do(X=x) denotes an intervention that sets the random variable(s) $X$ to $x$.
\item $\P{Y|do(X)}$ is the distribution of $Y$ conditional on an \emph{intervention} that sets $X$. This notation is somewhat overloaded. It may be used represent a probability, a probability distribution/mass function or a family of distribution functions depending on if the variables are discrete or continuous and whether or not we are treating them as fixed. For example it could represent 
\begin{itemize}
\item The probability $\P{Y=1|do(X=x)}$ as a function of $x$
\item The probability mass function for a discrete $Y$ : $\P{Y|do(X=x)}$
\item The probability density function for a continuous  $Y$ : $f_Y(y|do(X=x))$
\item a familiarly of density/mass function for $Y$ paramaterised by $x$.
\end{itemize}
Where the distinction is important not clear from context we will use one of the more specific forms above. 
\end{itemize}
\end{definition}

\begin{theorem}[Truncated product formula]
\label{thm:truncated_prodcut}
If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by simply dropping the terms for each of the variables in $Z$ from the factorisation given by the network \citep{Pearl2000}.

\begin{equation}
\label{eq:truncatedproduct}
\P{X_1...X_N|do(Z=z)} =
  \begin{cases}
  \prod_{i \notin Z}\P{X_{i}|parents_{X_{i}}} & \text{if $(X_{1}...X_{N})$ consistant with $Z=z$}  \\
   0       & \text{otherwise } 
  \end{cases}
\end{equation}
\end{theorem}

Theorem \ref{thm:truncated_prodcut} does not hold for standard bayesian networks for the simple reason that there are multiple valid networks for the same distribution. The truncated product formula will give different results depending on which you choose. The result is possible with causal bayesian networks because it follows directly from the assumption that the direction of the link indicates causality. In fact, from the interventionist viewpoint of causality, the truncation product formula defines what it means for a link to be causal.

Returning to example \ref{exm:ranomized_experiment}, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $\P{Y|do(X=1)}$, relative to if no one was treated $\P{Y|do(X=0)}$? The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and disease sub type of the patient. Since $X$ is assigned  via deliberate randomisation we know that it is not effected by any latent variables. The causal bayesian network for this scenario is shown in figure \ref{fig:causal_network_example} .This network represents the (causal) factorisation  $\P{X,Y} = \P{X}\P{Y|X}$, so from equation (\ref{eq:truncatedproduct}), $\P{Y|do(X)} = \P{Y|X}$. In this example, the interventional distribution is equivalent to the observational one.

@@

In example \ref{exm:adjusting} we are interested $\P{Y|do(X=1)}$, the expected high-school graduation rate if we introduce universal preschool. We could compare it to outlawing pre-school $\P{Y|do(X=0)}$ or the current status quo $\P{Y}$. It seems reasonalble to assume that attending could pre-school effect the likelyhood of graduating from high school and that your parents socio-economic status effects both pre-school attendance and high-school graduation. If we assume that socio-economic status is the only such variable (nothing else effects both attendance \emph{and} graduation), we can represent this problem with the causal bayesian network in figure \ref{exm:adjusting}. In this case, the interventional distribution is not equivelent to the observational one. If better off parents are more likely to send their children to pre-school these children more likely to graduate high school regardless, then simply comparing the high school graduation rates of those who attended preschool with those who did not will overstate the benifit of preschool. To obtain the interventional distribution we have to estimate the benifit of preschool for each socio-economic level seperately and then weight the results by the proportion of the population in that group,

@@

We have seen from these two examples that the expression to estimate the causal effect of an intervention depends on the structure of the causal graph. There is a very powerful and general set of rules that specify how we can transform observational distributions into intervetional ones for a given graph structure. These rules are referred to as the Do-calculus \cite{Pearl2000}. We discuss them further in section \ref{}.

\begin{figure}
\center
\caption{Causal bayesian network for example \ref{exm:adjusting}}
\label{fig:causal_adjust}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observedrect},lt/.style={latent}]

@@

Cyclic models \cite{Hyttinen2012}

@@

\section{Addressing discrimination}

>>>>

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    punkt/.style={
           circle,
           rounded corners,
           draw=black, thick,
           text width=1em,
           minimum height=1em,
           text centered},
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.3em,
           minimum height=2.3em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered
           },
    target/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
    observedrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    latentrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered
           },
     targetrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }   
}

\newcommand{\KL}{\operatorname{KL}}
\newcommand{\simpleregret}{R_T}
\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
\newcommand{\Q}[1]{\operatorname{Q}\left(#1\right)}
\newcommand{\EE}{\mathbb E}
\newcommand{\EEa}{\EE_a}
\newcommand{\Pns}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
\newcommand{\Pn}[2]{\operatorname{P}\left\{#2|#1\right\}}
\newcommand{\actions}{\mathcal{A}}
\newcommand{\calA}{\mathcal A}
\newcommand{\etc}{\textit{etc}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\calP}{\mathcal P}
\newcommand{\x}{\boldsymbol{x}}
\newcommand{\Ps}{\operatorname{P}}
\newcommand{\R}{\mathbb R}
\newcommand{\Pri}[1]{\operatorname{P}_i\left\{#1\right\}}
\newcommand{\Prz}[1]{\operatorname{P}_0\left\{#1\right\}}
\newcommand{\indep}[1]{\mathcal{I}\left(#1\right)}
\newcommand{\vb}[1]{\boldsymbol{#1}}
\newcommand{\card}[1]{\left\vert{#1}\right\vert}
\newcommand{\actionspace}{\mathcal{A}}

\newcommand{\Esub}[2]{\mathbb E_{#1}\left[{#2}\right]}
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}[1]{\mathbb E\left[{#1}\right]}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}

\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}

\renewcommand{\P}[1]{\operatorname{P}\left(#1\right)}
\newcommand{\cf}[2]{{#1}^{#2}}
\newcommand{\bernoulli}{\operatorname{Bernoulli}}
\newcommand{\dirac}{\operatorname{Dirac}}
\newcommand{\parents}[1]{\operatorname{\mathcal{P}a}_{#1}}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
\newcommand{\regret}{\bar{R}_{T}} %psuedo-regret
\newcommand{\expectedregret}{\mathbb{E}R_{T}} % expected regret
\newcommand{\rawregret}{R_{T}} % plain regret 
\newcommand{\quotes}[1]{``#1''}

@@

\chapter*{Todo}
\begin{enumerate}
\item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
\end{enumerate}

@@

The rise of big data sets and powerful computers has seen an explosion in the application of machine learning. From health care, to entertainment and self driving cars; machine learning algorithms will transform many industries. It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge data sets heralds the end of theory (Reference) and that we may be only a short step from the Singularity, where artificial intelligence exceeds our own and then grows exponentially.

@@

The notion of causality has been widely debated in science and philosophy \cite{} but is still widely viewed as poorly defined \cite{}. This has led to a reluctance among applied researchers in many fields to refer to causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}. However, the magnitude, direction and even existence of an association depends on what other variables we adjust for (or include in a regression). Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via common sense if the association reported is the \emph{right one}.

@@

I do not find this distinction useful. 
We can only change the future - history is useful only as far as it tells us something about the future.
Reverse causal questions can be reposed as forward ones, when making the translation a reverse causal query will be effectively asking about many possible interventions (rather than just one). Problems highlighted as intractable in the reverse causal sense are also intractable in the forward inference form, typically because concern situations for which we do not have a sufficient number of similar instances to allow statistical reasoning. For example, the war question posed in Gelman.

A distinction between forward causal inference, what happens if we do X and reverse casual inference

@@

\section{Identifying when we have a causal problem}

@@

What about predicting the risk of death in patients with pneumonia? Suppose we wish to use the model to decide who should be treated in hospital and who can be sent home with antibiotics. If we assume that in hospital treatment is more effective this seems like a straightforward prediction problem. It is not. Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if we start using the model to decide whom to admit. (xxx et al) found exactly this effect. Their model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pneumonia. They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk. There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original addition and treatment protocols. However, using it to decide on what basis to admit people could kill. The key is understanding exactly what question you are asking. In this case we are care about what happens to patients with characteristics X if we treat them according to decision rule Z.

@@

\item To decide between actions we only need to rank them (not estimate their actual effect). 
\item The predicted outcome in the absence of an intervention provides a single point. We can use this to find which problems are most serious if left alone - and prioritise those for modelling changes. 
\item Any decision we take does not significantly impact the system from which the data was drawn to make it (for repeat decision making)
\item Does acting on the result of the prediction change the predictive distribution p(y|x)? IE change people's behaviour.
\item Ethics - ... People's viewpoint on if its OK...
\end{itemize}

@@

Causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it. To do this we need terminology to describe actions and how we anticipate the system will respond to them. Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks. In this chapter we will examine the problems these approaches allow us to solve, the assumptions they rely on and how they differ. We will also use them to describe the following simplified examples. The aim is to demonstrate the notations and formalisms needed to tackle more interesting problems later on.

\vspace*{.3cm}
\begin{example}
\label{exm:ranomized_experiment}
Suppose a pharmaceutical company wants to assess the effectiveness of a new drug on recovery from a given illness. This is typically tested by taking a large group of representative patients and randomly assigning half of them to a treatment group (who receive the drug) and the other half to a control group (who receive a placebo). The goal is to determine the clinical impacts of the drug by comparing the differences between the outcomes for the two groups (in this case, simplified to only two outcomes - recovery or non-recovery). We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome. 
\end{example}

\vspace*{.3cm}
\begin{example}
\label{exm:adjusting}
Suppose we want to estimate the impact on high school graduation rates of compulsory preschool for all 4 year olds. We have a large cross-sectional data set on a group of 20 year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES). We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectiveness of early childhood education including a landmark randomised trial, the Perry Preschool project, which ran from 1962-1967 \citep{weikart1970longitudinal}.}
\end{example}

\section{Causal Bayesian networks}

@@

\begin{figure}[h]
\centering
\caption{A general Bayesian network for the joint distribution over three variables. This network does not encode any conditional independencies between its variables and can thus represent any distribution over three variables.}
\label{fig:bayesnet}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]

@@

\begin{figure}
\centering
\caption{Some valid Bayesian networks for a distribution that an be factorised as $\P{X_1,X_2,X_3} = \P{X_1}\P{X_2|X_1}\P{X_3|X_2}$ (which implies $X_3 \ci X_1 | X_2$)}
\label{fig:bayesnet2}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (3) edge (2)
    (2) edge (1);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, below left=of 1](2){$X_{1}$};
\node[main node, below right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, above left=of 1](2){$X_{1}$};
\node[main node, above right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (2) edge (1)
    (3) edge (1)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\end{figure}

The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorised over the nodes and edges in the graph. There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$, (the converse is not true $G$ can have extra edges). If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;

\eqn{
\P{X_{1},...,X_{N}} = \prod_{i = 1...N}\P{X_{i}|parents_{X_{i}}}
}

A causal Bayesian network is a Bayesian network in which a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$. This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change. We need some notation to describe interventions and represent distributions over variables in the network after an intervention. In this thesis I use the do operator introduced by Pearl \citep{Pearl2000}.

\begin{definition}{The do-notation}
\begin{itemize}
\item $do(X=x)$ denotes an intervention that sets the random variable(s) $X$ to $x$.
\item $\P{Y|do(X)}$ is the distribution of $Y$ conditional on an \emph{intervention} that sets $X$. This notation is somewhat overloaded. It may be used represent a probability distribution/mass function or a family of distribution functions depending on whether the variables are discrete or continuous and whether or not we are treating them as fixed. For example it could represent 
\begin{itemize}
\item the probability $\P{Y=1|do(X=x)}$ as a function of $x$,
\item the probability mass function for a discrete $Y$ : $\P{Y|do(X=x)}$,
\item the probability density function for a continuous  $Y$ : $f_Y(y|do(X=x))$,
\item a family of density/mass function for $Y$ parameterised by $x$.
\end{itemize}
Where the distinction is important and not clear from context we will use one of the more specific forms above. 
\end{itemize}
\end{definition}

\begin{theorem}[Truncated product formula \citep{Pearl2000}]
\label{thm:truncated_prodcut}
If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by dropping the terms for each of the variables in $Z$ from the factorisation given by the network.

\begin{equation}
\label{eq:truncatedproduct}
\P{X_1...X_N|do(Z=z)} =
  \begin{cases}
  \prod_{i \notin Z}\P{X_{i}|parents_{X_{i}}} & \text{if $(X_{1}...X_{N})$ consistent with $Z=z$}  \\
   0       & \text{otherwise } 
  \end{cases}
\end{equation}
\end{theorem}

Theorem \ref{thm:truncated_prodcut} does not hold for standard Bayesian networks because there are multiple valid networks for the same distribution. The truncated product formula will give different results depending on the selected network. The result is possible with causal Bayesian networks because it follows directly from the assumption that the direction of the link indicates causality. In fact, from the interventionist viewpoint of causality, the truncation product formula defines what it means for a link to be causal.

Returning to example \ref{exm:ranomized_experiment}, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $\P{Y|do(X=1)}$, relative to if no one was treated $\P{Y|do(X=0)}$? The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and the disease sub-type of the patient. Since $X$ is assigned via deliberate randomisation we know that it is not affected by any latent variables. The causal Bayesian network for this scenario is shown in figure \ref{fig:causal_network_example}. This network represents the (causal) factorisation  $\P{X,Y} = \P{X}\P{Y|X}$, so from equation (\ref{eq:truncatedproduct}), $\P{Y|do(X)} = \P{Y|X}$. In this example, the interventional distribution is equivalent to the observational one.

@@

In example \ref{exm:adjusting} we are interested in $\P{Y|do(X=1)}$, the expected high-school graduation rate if we introduce universal preschool. We could compare it to outlawing preschool $\P{Y|do(X=0)}$ or the current status quo $\P{Y}$. It seems reasonable to assume that preschool attendance affects the likelihood of high school graduation \footnote{The effect does not have to be homogeneous, it may depend non-linearly on characteristics of the child, family and school.} and that parental socio-economic status would affect \emph{both} the likelihood of preschool attendance and high school graduation. If we assume that socio-economic status is the only such variable (nothing else effects both attendance \emph{and} graduation), we can represent this problem with the causal Bayesian network in figure \ref{fig:causal_adjust}. In this case, the interventional distribution is not equivalent to the observational one. If parents with high socio-economic status are more likely to send their children to preschool and these children are more likely to graduate high school regardless, comparing the graduation rates of those who attended preschool with those who did not will overstate the benefit of preschool. To obtain the interventional distribution we have to estimate the impact of preschool on high school graduation for each socio-economic level separately and then weight the results by the proportion of the population in that group,

@@

We have seen from these two examples that the expression to estimate the causal effect of an intervention depends on the structure of the causal graph. There is a very powerful and general set of rules that specifies how we can transform observational distributions into interventional ones for a given graph structure. These rules are referred to as the Do-calculus \cite{Pearl2000}. We discuss them further in section \ref{sec:do_calculus}.

\begin{figure}
\center
\caption{Causal Bayesian network for example \ref{exm:adjusting}}
\label{fig:causal_adjust}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observedrect},lt/.style={latent}]

@@

%Cyclic models \cite{Hyttinen2012}

@@

%\section{Addressing discrimination}
